Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-29 paper/gd1.tex
long introduction

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Most stellar streams discovered around the Milky Way reside in the Galactic halo
and are therefore primarily sensitive to the dark matter distribution.
The longest and most prominent thin stream --- the GD-1 stream --- has been used
to constrain the shape of the inner halo, and is a promising place to search for
signatures of dark matter subhalo interactions.
% Aims
Here we present a sample of highly-probable members of GD-1, selected to have
small parallaxes and retrograde proper motions using data from the \gaia\
mission, and \pans\ photometry consistent with an old and metal-poor population
at a distance of $\sim10~\kpc$.
This selection extends the stream by $20^\circ$, reveals a plausible location of
the progenitor, and shows significant density variations along the stream in
unprecedented detail.
In addition to several prominent gaps, for the first time we also find evidence
of stream members off the main track.
Such features are a natural result of a stream interaction with a massive
perturber, or an indication of complex internal kinematics within the
stream progenitor.
If caused by a subhalo interaction, we estimate that the detected underdensity
and the associated spur could have been induced by a perturbation of a
$\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
\end{abstract}

@@

Over 30 candidate stellar streams have been discovered throughout the halo of
the Milky Way, thanks to large-area, multi-band photometric surveys (especially
the Sloan Digital Sky Survey, SDSS; \citealt{York:2000}).
The Milky Way streams have a wide range of properties such as length, velocity
dispersion, and density (e.g., \citealt{Odenkirchen:2001, Grillmair:2006,
Grillmair:2006b, Belokurov:2006, Belokurov:2007, Bonaca:2012, Shipp:2018}; see
also the recent review \citealt{Grillmair:2016, Newberg:2016}), and the majority
have no obvious, bound progenitor systems;
of the thin, likely globular-cluster-origin streams, only two have clear
progenitors (NGC 5466 and Palomar 5).

The most prominent and longest (in angular extent) thin stream is the GD-1
stream (named after its discoverers; \citealt{Grillmair:2006}).
GD-1 was found using a matched filter on photometric data from the SDSS and
was found to span $\approx 60^\circ$ in Heliocentric sky coordinates, partially
because of its relative proximity to the Sun ($\approx 7$--$10~\textrm{kpc}$).
Despite being close and relatively high surface-density, no remnant progenitor
cluster or stellar system has been found associated with the stream, but its
width,  metallicity ($\feh \approx -1.4$; \citealt{Koposov:2010}) and estimated
stellar mass ($M \approx 2 \times 10^4~\msun$; \citealt{Koposov:2010}) are
consistent with being a disrupted globular cluster.

The long physical length of the GD-1 stream ($\sim 15~\kpc$) and location in the
Galactic halo (pericentric distance $r_\textrm{peri} \sim 13~\kpc$) make it an
ideal object for constraining the gravitational potential of dark matter in the
inner Milky Way.
Indeed, both from fitting orbits to binned phase-space measurements along the
GD-1 stream (\citealt{Koposov:2010}) and from modeling the stream track in
phase-space coordinates \citep{Bowden:2015, Bovy:2016}, it has been used to show that the
dark matter distribution within its orbit is consistent with being close to
spherical.

Its long length --- and therefore its implied old dynamical age --- combined
with its particular orbit also make it a prime stream to search for interactions
with dark matter subhalos:
because of its large pericentric distance and retrograde orbit with respect to
the Galactic bar, density variations in the stream are not expected from either
interactions with giant molecular clouds (\citealt{Amorisco:2016}) or resonant
encounters with the bar (\citealt{Pearson:2017}).
By estimating a stream age of $\approx 4~\textrm{Gyr}$ assuming a progenitor
mass of $10^5~\msun$, and by taking a subhalo mass function and density from
XXX, \citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
($\approx 5$--$7^\circ$) gap caused by an interaction with a
$10^6$--$10^7~\msun$ subhalo.
\todo{implications of subhalo detection (cite bullock review along the lines: on small scales, lcdm has some issues, so alternative classes of models proposed, some that lack structure on small scales; so ruling on the existence of subhalos expected to be detectable with GD-1 directly informs the nature of dark matter.)}

\todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
(\citealt{Carlberg:2013}).
More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\circ$ span of the stream (\citealt{DeBoer:2018}):
[summarize their conclusions].
\todo{these analyses relied on binned data, still contaminated by MW stars}

In this \article, we present clearest view of the GD-1 stream to date using
astrometric data from the \gaia\ mission data release 2 (\DR) combined with
precise photometry from the \pans\ survey.
We find XX under-densities gaps, some consistent with ...
We revisit...do not try to model or fit for Galactic potential.
Focus on stream properties, data.

@@

@@

\begin{figure}[h]
\begin{center}
% \includegraphics[width=\textwidth]{spur.pdf}
\end{center}
\caption{%
\todo{APW}
\label{fig:orbit}
}
\end{figure}

@@

- future:
-- radial velocities, measure 3d velocity structure (unique predictions)
-- tests for other outstanding issues
-- model facets of stream morphology to ascertain if consistent with encounter signatures

>>>>

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Most stellar streams discovered around the Milky Way reside in the Galactic halo
and are therefore primarily sensitive to the dark matter distribution.
The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
to constrain the shape of the inner halo, and is a promising place to search for
signatures of dark matter subhalo interactions.
% Aims
Here we present a sample of highly-probable members of GD-1, selected to have
small parallaxes and retrograde proper motions using data from the \gaia\
mission, and \pans\ photometry consistent with an old and metal-poor population
at a distance of $\sim10~\kpc$.
This selection extends the stream by $20^\circ$, reveals a plausible location of
the progenitor, and shows significant density variations along the stream in
unprecedented detail.
In addition to several prominent gaps, for the first time we also find evidence
of stream members off the main track.
Such features are a natural result of a stream interaction with a massive
perturber, or an indication of complex internal kinematics within the
stream progenitor.
If caused by a subhalo interaction, we estimate that the detected underdensity
and the associated spur could have been induced by a perturbation of a
$\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
\end{abstract}

@@

Over 30 candidate stellar streams have been discovered throughout the halo of
the Milky Way, thanks to large-area, multi-band photometric surveys (especially
the Sloan Digital Sky Survey, SDSS; \citealt{York:2000}).
The Milky Way streams have a wide range of properties such as length, velocity
dispersion, and density (e.g., \citealt{Odenkirchen:2001, Grillmair:2006,
Grillmair:2006b, Belokurov:2006, Belokurov:2007, Bonaca:2012, Shipp:2018}; see
also the recent review \citealt{Grillmair:2016, Newberg:2016}).
% , and the majority
% have no obvious, bound progenitor systems;
% of the thin, likely globular-cluster-origin streams, only two have clear
% progenitors (NGC 5466 and Palomar 5).

The most prominent and longest (in angular extent) thin stream is the GD-1
stream (named after its discoverers; \citealt{Grillmair:2006}).
GD-1 was found using a matched filter on photometric data from the SDSS and
was found to span $\approx 60^\circ$ on the sky, partially
because of its relative proximity to the Sun ($\approx 7$--$10~\textrm{kpc}$).
Despite being close and relatively high surface-density, no remnant progenitor
cluster or stellar system has been found associated with the stream, but its
width ($\sigma \approx 12'$),  metallicity ($\feh \approx -1.4$) and estimated
stellar mass ($M \approx 2 \times 10^4~\msun$) are
consistent with being a disrupted globular cluster \citep{Koposov:2010}.

The long physical length of the GD-1 stream ($\sim 15~\kpc$) and location in the
Galactic halo (pericentric distance $r_\textrm{peri} \sim 13~\kpc$) make it an
ideal object for constraining the gravitational potential of dark matter in the
inner Milky Way.
Indeed, both from fitting orbits to binned phase-space measurements along the
GD-1 stream (\citealt{Koposov:2010}) and from modeling the stream track in
phase-space coordinates \citep{Bowden:2015, Bovy:2016}, we learned that the
dark matter distribution in the inner $\approx20\,\rm kpc$ is consistent with being close to
spherical.
This is in line with findings from simulated Milky Way-like galaxies; while dark matter halos are in general expected to be triaxial \citep{Allgood:2006}, the baryonic processes forming galaxies tend to make their inner regions more spherical \citep{Zhu:2016}.

An implied old dynamical age, combined with its orbital properties, makes GD-1 a prime stream to not only map the dark matter on large scales, but to also study its small scale structure.
Density variations in the stream can be induced from interactions with dark matter subhalos \citep[e.g.,][]{Ngan:2014}, but also from interactions with giant molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar \citep{Pearson:2017}.
However, these baryonic perturbations are unlikely in the case of GD-1 because of its large pericentric distance and retrograde orbit with respect to
the Galactic bar.
By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
mass of $10^5~\msun$, and by assuming properties of the subhalo population from dark matter-only simulations \citep{Springel:2008, Diemand:2008} (scaled to the Milky Way halo mass of $10^{12}\,\msun$, and reduced in number by a factor of 3 to account for disruption by the Galactic disk), \citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
($\approx 5$--$7^\circ$) gap caused by an interaction with a
$10^6$--$10^7~\msun$ subhalo.
A definitive ruling on the presence of dark matter subhalos in this mass regime would directly inform on the nature of dark matter \citep[e.g.][]{Bullock:2017}.

Photometric studies of GD-1 revealed the presence of density variations and gaps along the stream, as well as wiggles in the main stream track \citep[][]{Carlberg:2013, DeBoer:2018}.
Purely photometric stream studies suffer to some extent from contamination of the Milky Way foreground, whose small-scale spatial variations could partially account for density inhomogeneities reported along cold streams \citep[e.g.,][]{Ibata:2016}.
In this \article, we improve upon the selection of likely GD-1 members by using astrometric data from the \gaia\ mission data release 2 in addition to precise photometry from the \pans\ survey (\S\,\ref{sec:data}).
This data enabled us to measure global properties of GD-1 to an unprecedented precision (\S\,\ref{sec:res_global}), and detect clear signatures of stream stars beyond the main track (\S\,\ref{sec:res_gap}).
Implications of this revolutionary remapping of GD-1 are discussed in \S\,\ref{sec:discussion}.
% \todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
% (\citealt{Carlberg:2013}).
% More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
% interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\circ$ span of the stream (\citealt{DeBoer:2018}):
% [summarize their conclusions].
% \todo{these analyses relied on binned data, still contaminated by MW stars}
% 
% In this \article, we present clearest view of the GD-1 stream to date using
% astrometric data from the \gaia\ mission data release 2 (\DR) combined with
% precise photometry from the \pans\ survey.
% We find XX under-densities gaps, some consistent with ...
% We revisit...do not try to model or fit for Galactic potential.
% Focus on stream properties, data.

@@

- Spur vs. blob
- Estimate velocity offset and significance
- No indication of these features in previous work

@@

@@

- future:
-- radial velocities, measure 3d velocity structure (unique predictions)
-- tests for other outstanding issues
-- model facets of stream morphology to ascertain if consistent with encounter signatures
-- so far a lot of focus on gaps produced in encounters -- extra features we detected beside the GD-1 track suggest that further theoretical focus on scattered material is needed to fully understand its dynamical history
